---
title: "Assignment 4 - HPC and SQL"
author: "Xinran Wang"
link-citations: true
toc: false
editor: 
  markdown: 
    wrap: 72
fig-width: 8
fig-height: 6
theme: cosmos
---

## Due Date

This assignment is due by 11:59pm Pacific Time on Friday, November 21st,
2025.

The learning objectives are to write faster code for a computational
task requiring a loop and to implement some queries and basic data
wrangling in SQL.

```{r}
library(RSQLite)
library(DBI)
library(collapse)
library(parallel)
```

## HPC

### Make things run faster

Rewrite the following R functions to make them faster. It is OK (and
recommended) to take a look at StackOverflow and Google.

-   `fun1`: Find the total for each row of a numeric matrix. Output
    should be a vector with length equal to the number of rows of the
    input matrix.
-   `fun2`: Calculate the running (cumulative) total along each row of a
    numeric matrix. Output should be a matrix with dimensions equal to
    the input matrix.

```{r, eval = TRUE}
# Total for each row
fun1 <- function(mat) {
  n <- nrow(mat)
  ans <- double(n) 
  for (i in 1:n) {
    ans[i] <- sum(mat[i, ])
  }
  ans
}

# https://stackoverflow.com/questions/53772516/sum-by-row-dataframe-r-fastest-way
fun1alt <- function(mat) {
  rowSums(mat, na.rm = T)
}

# running cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

# https://stackoverflow.com/questions/7550383/calculating-cumulative-sum-for-each-row
# https://rdrr.io/cran/collapse/man/fcumsum.html
fun2alt <- function(mat) {
  t(apply(mat, 1, cumsum))
  # slower: t(apply(as.matrix(mat), 1, fcumsum)) 
}

```

### Question 1 (30 points)

Using the dataset generated below (`dat`), check that both of your new
functions produce the same outputs as the corresponding original
functions.

```{r, eval = TRUE}
# Use the data with this code
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)
```

Then use `microbenchmark` (as in the code below) to check that your
version is actually faster. How much faster is it?

```{r, eval = TRUE}
# Test for the first
microbenchmark::microbenchmark(
  fun1(dat),
  fun1alt(dat), unit = "relative"
)

identical(fun1(dat), fun1alt(dat))
```

[fun1alt is \~10 times faster than fun1 on average.]{style="color:blue"}

```{r, eval = TRUE}
# Test for the second
microbenchmark::microbenchmark(
  fun2(dat),
  fun2alt(dat), unit = "relative"
)

identical(fun2(dat), fun2alt(dat))
```

[fun2alt is \~2.6 times faster than fun2 on
average.]{style="color:blue"}

### Make things run faster with parallel computing

The following function allows us to estimate the value of $\pi$ through
simulation:

```{r, eval = TRUE}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132
```

In order to get accurate estimates, we can run this function multiple
times, with the following code:

```{r, eval = TRUE}
# This runs the simulation a 4,000 times, each with 10,000 points
set.seed(1231)
system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans))
})
```

### Question 2 (30 points)

Rewrite the previous code using `parLapply()` (or your parallelization
method of choice) to parallelize it. Run the code once, using
`system.time()`, to show that your version is faster.

```{r, eval = TRUE}
ncpus <- max(1, detectCores() - 1)
cl <- makeCluster(ncpus)
clusterSetRNGStream(cl, 123)
clusterExport(cl, c("sim_pi"), envir = environment())

set.seed(1231)
system.time({
  # ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  ans <- unlist(parLapply(cl, 1:4000, sim_pi, n = 10000))
  print(mean(ans))
})

stopCluster(cl)
```

## SQL

Setup a temporary database of movie data by running the following code
chunk:

```{r, eval = TRUE}
library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
film <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film.csv")
film_category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film_category.csv")
category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/category.csv")

# Copy data.frames to database
dbWriteTable(con, "film", film)
dbWriteTable(con, "film_category", film_category)
dbWriteTable(con, "category", category)
```

When you write a new chunk, remember to replace the `r` with
`sql, connection=con`. Some of these questions will require you to use
an inner join. Read more about them here
<https://www.w3schools.com/sql/sql_join_inner.asp>

```{r, eval = FALSE}
SELECT name, type
FROM sqlite_master;
```

```{r, eval = FALSE}
SELECT *
FROM film
LIMIT 5;
```

```{r, eval = FALSE}
SELECT *
FROM film_category
LIMIT 5;
```

```{r, eval = FALSE}
SELECT *
FROM category
LIMIT 5;
```

## Question 3 (10 points)

How many many movies are available in each `rating` category?

```{sql, connection=con}
SELECT 
    rating, 
    COUNT(film_id) AS num_films
FROM film
GROUP BY rating
ORDER BY num_films DESC;
```

## Question 4 (10 points)

What is the average replacement cost and rental rate for each `rating`
category?

```{sql, connection=con}
SELECT 
    rating,
    ROUND(AVG(replacement_cost), 2) AS avg_replacement_cost,
    ROUND(AVG(rental_rate), 2) AS avg_rental_rate
FROM film
GROUP BY rating
ORDER BY avg_rental_rate DESC;
```

## Question 5 (10 points)

Use table `film_category` together with `film` to find how many films
there are with each category ID.

```{sql, connection=con, max.print=50}
SELECT 
    film_category.category_id,
    COUNT(*) AS num_films
FROM film_category
INNER JOIN film
    ON film_category.film_id = film.film_id
GROUP BY film_category.category_id
ORDER BY num_films DESC
LIMIT 16;
```

## Question 6 (10 points)

Incorporate the `category` table into the answer to the previous
question to find the name of the most popular category.

```{sql, connection=con, max.print=50}
SELECT 
    category.name,
    film_category.category_id,
    COUNT(*) AS num_films
FROM film_category
INNER JOIN film
    ON film_category.film_id = film.film_id
INNER JOIN category
    ON film_category.category_id = category.category_id
GROUP BY category.name
ORDER BY num_films DESC
LIMIT 16;
```
