---
title: "Lab6"
author: "Xinran Wang"
format: html
editor: visual
---

# **Lab 6 - Text Mining**

```{r}
# knitr::opts_chunk$set(eval = FALSE, include  = TRUE)
```

# **Learning goals**

-   Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and n-grams from text.

-   Use dplyr and ggplot2 to analyze text data

# **Lab description**

For this lab we will be working with a new text-based dataset consisting of transcribed medical reports. The dataset contains transcription samples from <https://www.mtsamples.com/>. We have created a (somewhat) cleaned version of the data, which can be found at <https://raw.githubusercontent.com/USCbiostats/data-science-data/master/00_mtsamples/mtsamples.csv>.

### **Setup packages**

You should load in `dplyr`, `ggplot2` and `tidytext`. If you don’t already have `tidytext` then you can install with

```{r}
# install.packages("tidytext")
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(tidytext))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(textdata))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(ggplot2))
```

### **Read in Medical Transcriptions**

Loading in the cleaned data from the `USCbiostats/data-science-data` repository:

```{r}
mt_samples <- read_csv("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/00_mtsamples/mtsamples.csv")

mt_samples <- mt_samples |>
  select(description, medical_specialty, transcription)

head(mt_samples, 3)
```

------------------------------------------------------------------------

## **Question 1: What specialties do we have?**

Use the `count()` function from `dplyr` to figure out how many different categories we have in the data. Are these categories related? Overlapping? Evenly distributed?

```{r}
mt_samples_count <- mt_samples %>%
        count(medical_specialty, sort = TRUE) %>%
        rename(`Medical Specialty` = medical_specialty, 
               Count = n) %>%
        arrange(`Medical Specialty`)
```

```{r}
n <- nrow(mt_samples_count)
half <- ceiling(n / 2)

kable(data.frame(
  `Medical Specialty` = mt_samples_count$`Medical Specialty`[1:half],
  Count = mt_samples_count$Count[1:half],
  `Medical Specialty ` = mt_samples_count$`Medical Specialty`[(half+1):n],
  `Count ` = mt_samples_count$Count[(half+1):n]
), align = "c")
```

```{r}
ggplot(mt_samples_count, 
       aes(x = factor(`Medical Specialty`, 
                      levels = rev(`Medical Specialty`)), 
           y = Count)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Count of Medical Specialties",
    x = "Count",
    y = "Medical Specialty"
  ) +
  theme_minimal() 

rm(mt_samples_count)
```

------------------------------------------------------------------------

## **Question 2**

-   Tokenize the the words in the `transcription` column

-   Count the number of times each token appears

-   Visualize the top 20 most frequent words

Explain what we see from this result. Does it makes sense? What insights (if any) do we get?

[**Description.** The top 20 tokens are mostly function words (the, as, of, to, in), verbs (is, was, were), pronounces (he, she, patient). It makes sense in clinical transcripts.]{style="color:blue"}

```{r}
mt_samples_token <- mt_samples %>% 
  unnest_tokens(token, transcription)
```

```{r}
ggplot(mt_samples_token %>%
         count(token, sort = TRUE) %>% 
         slice_head(n = 20), 
       aes(x = reorder(token, -n), 
           y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Count of Top 20 Tokens Before Removing Stop Words",
    x = "Tokens",
    y = "Count"
  ) +
  theme_minimal() 
```

------------------------------------------------------------------------

## **Question 3**

-   Re-do the visualization but remove stop words before making it

-   Bonus points if you remove numbers as well

What do we see know that we have removed stop words? Does it give us a better idea of what the text is about?

[**Description.** After removing stop words, we can see more meaningful context (such as patient, left, history, normal, procedure, pain).]{style="color:blue"}

```{r}
kable(head(mt_samples_token %>%
             count(token, sort = TRUE), 5), 
        caption = "Top 5 tokens BEFORE removing stop word")

mt_samples_token <- mt_samples_token %>% 
  anti_join(stop_words, by = c("token" = "word")) %>%
  filter(!grepl("^[0-9]+$", token))

kable(head(mt_samples_token %>% 
             count(token, sort = TRUE), 5),
      caption = "Top 5 tokens AFTER removing stop word", 
      align = "c" )
```

```{r}
ggplot(mt_samples_token %>%
         count(token, sort = TRUE) %>% 
         slice_head(n = 20), 
       aes(x = reorder(token, -n), 
           y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Count of Top 20 Tokens After Removing Stop Words",
    x = "Tokens",
    y = "Count"
  ) +
  theme_minimal() 
```

------------------------------------------------------------------------

# **Question 4**

Repeat question 2, but this time tokenize into bi-grams. How does the result change if you look at tri-grams?

```{r}
mt_samples_token_bigram <- mt_samples %>% 
  unnest_ngrams(ngram, transcription, n = 2)

mt_samples_token_trigram <- mt_samples %>% 
  unnest_ngrams(ngram, transcription, n = 3)
```

```{r}
ggplot(mt_samples_token_bigram %>%
         count(ngram, sort = TRUE) %>% 
         slice_head(n = 20), 
       aes(x = reorder(ngram, -n), 
           y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Count of Top 20 Tokens (bi-gram)",
    x = "Tokens",
    y = "Count"
  ) +
  theme_minimal() 
```

```{r}
ggplot(mt_samples_token_trigram %>%
         count(ngram, sort = TRUE) %>% 
         slice_head(n = 20), 
       aes(x = reorder(ngram, -n), 
           y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Count of Top 20 Tokens (tri-gram)",
    x = "Tokens",
    y = "Count"
  ) +
  theme_minimal() 
```

------------------------------------------------------------------------

# **Question 5**

Using the results you got from Question 4, pick a word and count the words that appear before and after it.

```{r}
mt_samples_token_trigram_patient <- mt_samples_token_trigram %>% 
  separate(ngram, into = c("Word Before", "Patient", "Word After"), sep = " ") %>% 
  filter(Patient == "patient")
```

```{r}
kable(bind_cols(head(mt_samples_token_trigram_patient %>%
                       count(`Word Before`, sort = TRUE), 5) %>%
                  rename(`Count Before` = n), 
                head(mt_samples_token_trigram_patient %>%
                       count(`Word After`, sort = TRUE), 5) %>%
                  rename(`Count After` = n)), 
      caption = "Top 5 Words Before and After `Patient`", 
      align = "c" )
```

```{r}
rm(mt_samples_token_bigram, mt_samples_token_trigram, mt_samples_token_trigram_patient)
```

------------------------------------------------------------------------

# **Question 6**

Which words are most used in each of the specialties? You can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty. Remember to remove stop words. What are the 5 most-used words for each specialty?

```{r}
kable(
  mt_samples_token %>% 
    group_by(medical_specialty) %>%
    count(token, sort = TRUE) %>%
    top_n(1) %>%
  rename(`Medical Specialty` = medical_specialty, 
         Word = token, 
         Count = n),
  caption = "Most Used Words by Medical Specialties", 
  align = "c")
```

# **Question 7 - extra**

Find your own insight in the data:

Ideas:

-   Interesting n-grams

-   See if certain words are used more in some specialties than others

```{r}
mt_samples_tf_idf <- mt_samples %>%
  unnest_tokens(transcription, transcription) %>%
  anti_join(stop_words, by = c("transcription" = "word")) %>%
  count(transcription, medical_specialty) %>%
  bind_tf_idf(transcription, medical_specialty, n)
```

```{r}
kable(mt_samples_tf_idf %>% 
        group_by(medical_specialty) %>%
        arrange(desc(tf)) %>%
        top_n(1) %>%
        select(transcription, medical_specialty, n, tf) %>%
        arrange(medical_specialty) %>%
        rename(Transcription = transcription,
               `Medical Speciality` = medical_specialty), 
      caption = "Most Frequent Term by Medical Specialty (tf)", align = "c")
```

```{r}
kable(mt_samples_tf_idf %>% 
        arrange(desc(idf)) %>%
        top_n(10) %>%
        select(transcription, medical_specialty, n, idf) %>%
        rename(Transcription = transcription,
               `Medical Speciality` = medical_specialty), 
      caption = "10 Rarest Term by Medical Specialty (idf)", 
      align = "c")
```

```{r}
kable(mt_samples_tf_idf %>% 
        group_by(medical_specialty) %>%
        arrange(desc(tf_idf)) %>%
        top_n(1) %>%
        select(transcription, medical_specialty, n, tf_idf) %>%
        arrange(medical_specialty) %>%
        rename(Transcription = transcription,
               `Medical Speciality` = medical_specialty), 
      caption = "Most Distinctive Term by Medical Specialty (tf_idf)", 
      align = "c")
```
