---
title: "Lab9"
author: "Xinran Wang"
format: html
editor: visual
---

## **Lab 9 - SQL**

# **Setup**

```{r}
# install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
actor <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/actor.csv")
rental <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/rental.csv")
customer <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/customer.csv")
payment <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/payment_p2007_01.csv")

# Copy data.frames to database
dbWriteTable(con, "actor", actor)
dbWriteTable(con, "rental", rental)
dbWriteTable(con, "customer", customer)
dbWriteTable(con, "payment", payment)
```

```{r}
dbListTables(con)
```

TIP: You can use the following QUERY to see the structure of a table

```{r}
#PRAGMA table_info(actor)
dbGetQuery(con, "PRAGMA table_info(actor);")
#dbGetQuery(con, "PRAGMA table_info(customer);")
#dbGetQuery(con, "PRAGMA table_info(payment);")
#dbGetQuery(con, "PRAGMA table_info(rental);")
```

SQL references:

<https://www.w3schools.com/sql/>

```{r}
colnames(dbGetQuery(con, "PRAGMA table_info(actor);"))
#colnames(dbGetQuery(con, "PRAGMA table_info(customer);"))
#colnames(dbGetQuery(con, "PRAGMA table_info(payment);"))
#colnames(dbGetQuery(con, "PRAGMA table_info(rental);"))
```

# **Exercise 1**

Edit the code below to retrieve the actor ID, first name and last name for all actors using the `actor` table. Sort by last name and then by first name (note that the code chunk below is set up to run SQL code rather than R code).

```{r}
head(dbGetQuery(con, "
  SELECT actor_id, first_name, last_name
  FROM actor
  ORDER BY last_name, first_name;
"))
```

# **Exercise 2**

Retrieve the actor ID, first name, and last name for actors whose last name equals ‘WILLIAMS’ or ‘DAVIS’.

```{r}
dbGetQuery(con, "
  SELECT actor_id, first_name, last_name
  FROM actor
  WHERE last_name = 'WILLIAMS' OR last_name = 'DAVIS'
  ORDER BY last_name, first_name;;
")
```

# **Exercise 3**

Write a query against the `rental` table that returns the IDs of the customers who rented a film on July 5, 2005 (use the rental.rental_date column, and you can use the `date()` function to ignore the time component). Include a single row for each distinct customer ID.

```{r}
head(dbGetQuery(con, "
  SELECT DISTINCT customer_id
  FROM rental
  WHERE date(rental_date) = '2005-07-05';
"))
```

# **Exercise 4**

## **Exercise 4.1**

Construct a query that retrieves all rows from the `payment` table where the amount is either 1.99, 7.99, 9.99.

```{r}
head(dbGetQuery(con, "
  SELECT *
  FROM payment
  WHERE amount IN (1.99, 7.99, 9.99);
"))
```

## **Exercise 4.2**

Construct a query that retrieves all rows from the `payment` table where the amount is greater then 5.

```{r}
head(dbGetQuery(con, "
  SELECT *
  FROM payment
  WHERE amount > 5;
"))
```

## **Exercise 4.2**

Construct a query that retrieves all rows from the `payment` table where the amount is greater then 5 and less then 8.

```{r}
head(dbGetQuery(con, "
  SELECT *
  FROM payment
  WHERE amount > 5 AND amount < 8;
"))
```

# **Exercise 5**

Retrieve all the payment IDs and their amounts from the customers whose last name is ‘DAVIS’.

```{r}
dbGetQuery(con, "
  SELECT payment.payment_id, payment.amount
  FROM payment
  INNER JOIN customer
    ON payment.customer_id = customer.customer_id
  WHERE customer.last_name = 'DAVIS';
")
```

# **Exercise 6**

## **Exercise 6.1**

Use `COUNT(*)` to count the number of rows in `rental`.

```{r}
dbGetQuery(con, "
  SELECT COUNT(*)
  FROM rental;
")
```

## **Exercise 6.2**

Use `COUNT(*)` and `GROUP BY` to count the number of rentals for each `customer_id`.

```{r}
head(dbGetQuery(con, "
  SELECT customer_id, COUNT(*)
  FROM rental
  GROUP BY customer_id;
"))
```

## **Exercise 6.3**

Repeat the previous query and sort by the count in descending order.

```{r}
head(dbGetQuery(con, "
  SELECT customer_id, COUNT(*) AS rental_count
  FROM rental
  GROUP BY customer_id
  ORDER BY rental_count DESC;
"))
```

## **Exercise 6.4**

Repeat the previous query but use `HAVING` to only keep the groups with 40 or more.

```{r}
dbGetQuery(con, "
  SELECT customer_id, COUNT(*) AS rental_count
  FROM rental
  GROUP BY customer_id
  HAVING COUNT(*) >= 40
  ORDER BY rental_count DESC;
")
```

# **Exercise 7**

Write a query that calculates a number of summary statistics for the payment table using `MAX`, `MIN`, `AVG` and `SUM`

```{r}
dbGetQuery(con, "
  SELECT 
    MAX(amount),
    MIN(amount),
    AVG(amount),
    SUM(amount) 
  FROM payment;
")
```

## **Exercise 7.1**

Modify the above query to do those calculations for each `customer_id`.

```{r}
head(dbGetQuery(con, "
  SELECT 
    customer_id,
    MAX(amount),
    MIN(amount),
    AVG(amount),
    SUM(amount) 
  FROM payment
  GROUP BY customer_id;
"))
```

## **Exercise 7.2**

Modify the above query to only keep the `customer_id`s that have more then 5 payments.

```{r}
head(dbGetQuery(con, "
  SELECT 
    customer_id,
    MAX(amount),
    MIN(amount),
    AVG(amount),
    SUM(amount) 
  FROM payment
  GROUP BY customer_id;
  HAVING COUNT(*) > 5;
"))
```

# **Cleanup**

Run the following chunk to disconnect from the connection.

```{r}
# clean up
dbDisconnect(con)
```
